Human Factors Influencing Visual Statistical Inference

### 2013-06-10 sociology_paper.tex
overall edits, giving reference to figures

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement in the lineup does not affect the inference.

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis. 
, indicating that we have enough evidence to reject the null hypothesis.
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer.
 \hh{Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.}
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

A brief description of the factors that may affect the power of visual inference.

\dc{You need to have more than one sentence to open this section, it needs a paragraph overviewing the factors.}

@@

\dc{Learning trend and location effect need to be separate sections, cannot be simply bolded.}

@@

\paragraph{Learning Trend:} Learning trend of an observer can be observed in terms of performance over successive feedback when multiple lineups are shown for evaluation. Experiments 5, 6 and 7 were used for this. Each subject was shown a total of 10 lineups randomly selected from a pool of lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups is randomized. The responses of the lineups are recorded by attempt 1 through 10. Attempt 1 means the response is for the first lineup the observer evaluates and attempt 10 refers to the response for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\paragraph{Design for Location effect:} The experiment was set up based on the gene expression data with two groups. The groups are defined based on the Genotype and Empty Vector (EV). For each of the of the Genotype, gene expression data were collected in presence or absence of iron sufficiency. Two hypothesis were of primary interest. The Genotype effect and the Genotype and iron sufficiency interaction effect.

@@

Besides United States and India, countries such as Canada, Romania, United Kingdom and Macedonia have more than 10 participants each. The rest of the 70 countries have less than 10 participants.

@@

A total of XX lineups were evaluated in all these 10 experiments. Each person was supposed to evaluate 10 lineups except experiment 9 where three lineups were evaluated by each person. To clean the data we used the test plot as the basis and did not include the results of that test plot. In some cases the demographic information were not provided by the participants. Also, for some ip address, the actual geographical locations could not be retrieved. This resulted some missing demographic information.

@@

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_response} fitted for \hh{probability of correct lineup evaluation.} %log time taken to evaluate a lineup. 
None of the fixed factor effects of attempt ($\alpha_2$ through $\alpha_{10}$) seem to be significantly different from %each other 
\hh{the first attempt $\alpha_1$}
at \%1 level in all three experiments 5, 6 and 7. For experiment 7 subject specific variation is very small on the other hand lineup variance is much higher compared to the other two experiments.}
\scalebox{0.9}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

With controlled simulation experiments, it is observed that the performance of human observers does not increase through successive attempts while evaluating multiple lineups. But their skill in evaluating the lineup in shorter time gets improved over successive evaluations. When an observer evaluates multiple lineups, the earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

Observer demographics don't seem to have significant influence on performance in time taken to evaluate a lineup. Some variation is observed in terms of median proportion correct, but the mean proportion correct over different demographic factors levels seem to be similar.

@@

More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of knowledge about statistical graphics. The subjects participating this study may not necessarily knowledgable about statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference may be higher with knowledgeable observer. Thus the variations observed with individual performances could be attributed to individual abilities in terms of their training on statistical graphics.

>>>>

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement of data plot in the lineup does not affect the inference.

@@

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis. 
, indicating that we have enough evidence to reject the null hypothesis.
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. 
(Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect-1} shows an `electoral building' where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generated as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true the actual data plot should look just like any of the other plots, and not be identified by an observer. Figure~\ref{fig:tower} shows a plot of the electoral building with added context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

@@

Based on human evaluation on a lineup a decision is made in visual statistical inference. While the visual test statistic is defined to have a direct influence on the observer to pick a plot based on the signal in the data, there are other human factors that may affect the observer performance. It is important to study the effect of those factors and examine the extent of their influence. This section presents a brief description of the factors that may affect the power of visual inference.

%\dc{You need to have more than one sentence to open this section, it needs a paragraph overviewing the factors.}

@@

%\dc{Learning trend and location effect need to be separate sections, cannot be simply bolded.}

@@

\subsubsection{Learning Trend} Learning trend of an observer can be observed in terms of performance over successive feedback when multiple lineups are shown for evaluation. Experiments 5, 6 and 7 were used for this. Each subject was shown a total of 10 lineups randomly selected from a pool of lineups. The lineups are not necessarily with the same difficulty levels. But the order of lineups is randomized. The responses of the lineups are recorded by attempt 1 through 10. Attempt 1 means the response is for the first lineup the observer evaluates and attempt 10 refers to the response for the 10th lineup. The goal is to estimate whether performance of the observer improves from attempt 1 to attempt 10.

\subsubsection{Design for Location effect} The experiment was set up based on the gene expression data with two groups. The groups are defined based on the Genotype and Empty Vector (EV). For each of the of the Genotype, gene expression data were collected in presence or absence of iron sufficiency. Two hypothesis were of primary interest. The Genotype effect and the Genotype and iron sufficiency interaction effect.

@@

Besides United States and India, countries such as Canada, Romania, United Kingdom and Macedonia have more than 10 participants each. The rest of the 70 countries have less than 10 participants. Figure \ref{fig:turker_location_experiment} displays the location of turk workers by each experiment. Notice that For all the ten experiments the locations are similar with some small variations. That may be due to the time of the experiment when it is started. For some experiments India got more participants than United States and for some experiments the numbers just got reversed (\ref{fig:participation_time}). It is also observed in some experiments number of  participants are similar no matter whether the experiment is run on day time or nigh time.

@@

A total of 1911 lineups were evaluated in all these 10 experiments. Each person was supposed to evaluate 10 lineups except experiment 9 where three lineups were evaluated by each person. To clean the data we used the test plot as the basis and did not include the results of that test plot. In some cases the demographic information were not provided by the participants. Also, for some ip address, the actual geographical locations could not be retrieved. This resulted some missing demographic information.

@@

\begin{table}[htbp]
\centering
\caption{Parameter estimates of Model \eqref{eqn:trend_response} fitted for probability of correct lineup evaluation. None of the fixed factor effects of attempt ($\alpha_2$ through $\alpha_{10}$) seem to be significantly different from the first attempt $\alpha_1$
at \%1 level in all three experiments 5, 6 and 7. For experiment 7 subject specific variation is very small on the other hand lineup variance is much higher compared to the other two experiments.}
\scalebox{0.9}{
\begin{tabular}{rrrrccrrrccrrrc}
  \hline
& \multicolumn{4}{c} {Experiment 5} & &\multicolumn{4}{c} {Experiment 6} && \multicolumn{4}{c} {Experiment 7}\\

@@

Observer demographics don't seem to have significant influence on performance in time taken to evaluate a lineup. Some variation is observed in terms of median proportion correct, but the overall average proportion of correct responses over different demographic factors levels seem to be similar.

With simulation experiments, it is observed that the performance of human observers does not increase through successive attempts while evaluating multiple lineups. But their skill in evaluating the lineup in shorter time gets improved over successive evaluations. When an observer evaluates multiple lineups, the earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

@@

The subjects participating this study may not necessarily know about statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference may be higher with knowledgeable observer. Thus the variations observed with individual performances could be attributed to individual abilities in terms of their training on statistical graphics. More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of knowledge about statistical graphics.
